\large{\bf Looking forwards and backwards through locally regulated populations }

### 2023-04-13 forwards-and-backwards.tex
first integral in the appendix done

\begin{lemma}
    \label{def: MP definition of limit}
For $G \in \mathcal{C}^{\infty}(\IR)$ with $\|G'''\|_\infty<\infty$,
and 
$f \in \mathcal{C}_{b}^\infty(\overline{\IR}^d)$, define the 
function $G_f$ by 
$G_f(\eta):=G (\langle f, \eta \rangle)$.
Let $\Pgen^N$ be the generator given by
\begin{equation}
\begin{aligned}
    \Pgen^N G_f(\eta)
    &:=
    \theta \bigg\langle
    \gamma(x, \eta)
        \int 
        \left(G(\langle f, \eta \rangle + f(z)/N) - G(\langle f, \eta \rangle)\right)
        r(z,\eta) q_\theta(x,dz)
    +
    \left(G(\langle f, \eta \rangle - f(x)/N) - G(\langle f, \eta \rangle)\right)
    \mu_\theta(x, \eta),
    \eta(dx) \bigg\rangle .
\end{aligned}
\end{equation}
A process $(\eta^N_t)_{t \geq 0}$ 
satisfies the martingale characterization of Definition~\ref{defn:mgale_construction}
if and only if it is a solution 
to the $(\Pgen^N, \eta_0)$-martingale problem, i.e.,
if for all such test functions 
$$
    M_t :=
    G_f(\eta^N_t) - G_f(\eta^N_0)
    - \int_{0}^{t}\Pgen^N G_f(\eta^N_s)ds
$$
is a martingale (with respect to the natural $\sigma$-field).

@@

To complete the description of the process,
each line of descent
gives birth to lines at higher levels at rate $2 (N - u) c_\theta(x, \eta)$,
and each such new line chooses a level uniformly from $[u, N]$,
a spatial location $y$ from the kernel
$$
    q^m(x, dy, \eta)
    =
    r(y, \eta) q(x, dy) / \int_{\IR^d} r(y, \eta) q(x, dy),
$$
and the two lines swap spatial locations with probability 1/2.

@@

\begin{proof}[of Lemma \ref{lem:test_fn_tightness}]
    Again, we use the Aldous-Rebolledo criterion.
    Tightness of $\langle f, \lp_t \rangle$ for a fixed $t$
    follows from Lemma~\ref{lem:compact containment hn},
    so we need only prove conditions analogous to
    \eqref{eqn:eta_projections_goal1}
    and \eqref{eqn:eta_projections_goal2}
    applied to the martingale representation
    of equations~\eqref{eqn:f_xi_mgale} and~\eqref{eqn:f_xi_qv}.
    Rewriting~\eqref{eqn:f_xi_mgale} with $c_\theta = c_\theta(x, \eta_s)$,
    \begin{align*}
        \langle f, \lp_t \rangle
        &=
        \langle f, \lp_0 \rangle
        + M^f_t
        + \int_0^t \Big\langle
            c_\theta \int_u^N \int (f(y, u_1) + f(x, u_1)) q^m(x, dy) du_1
        \\ {} &\qquad
            + c_\theta (N - u) \int (f(y, u) - f(x, u)) q^m(x, dy)
            + (c_\theta u^2 - b_\theta u) \frac{d}{du} f,
            \lp_s
        \Big\rangle ds
        \\ &\le
        \langle f, \lp_0 \rangle
        + M^f_t
        + \int_0^t \Big\langle
            c_\theta \int_u^N \int (f(y, u_1) + f(x, u_1)) q^m(x, dy) du_1
        \\ {} &\qquad
            + c_\theta (N - u) \int (f(y, u) - f(x, u)) q^m(x, dy)
            + (c_\theta u^2 - b_\theta u) \frac{d}{du} f,
            \lp_s
        \Big\rangle ds
    \end{align*}
    The bounds analogous to \eqref{eqn:eta_projections_goal1} and \eqref{eqn:eta_projections_goal2}
    follow as in the proof of Lemma~\ref{lem:eta_projections_tightness}:
    for instance, observe that using that $c_\theta \le C_0$ for some $C_0$,
    the predictable part of this semimartingale decomposition is bounded by
    \begin{align*}
        \Big\langle
            2 C_0 \|f\|_\infty u_f
            + (1 - u/N) \gamma B^\theta_f
            + (C_0 u^2 - b_\theta u) \frac{d}{du} f,
            \lp_s
        \Big\rangle
        &\le
        \langle
            C_0 u_f^2 + \sup_x |b_\theta(x, \eta_s)| u_f \|\frac{d}{du} f\|_\infty
        \rangle ,
    \end{align*}
    which can be bounded as we did for \eqref{eqn:eta_projections_goal1}.
\end{proof}

@@

In this section we verify one of the conditions of the Markov Mapping Theorem,
namely, that ``integrating out levels'' in the generator of the lookdown process
we obtain the generator of the projected process.
We will work with test functions of the form
$$  f(\lp) = \prod_{(x, u) \in \lp} g(x, u) = \exp\left(\langle \log g, \lp \rangle \right) . $$
Furthermore, recall that $\kappa^N(\lp)(\cdot) = \lp(\cdot \times [0, N)) / N$
is the ``spatial projection operator'',
and define the probability kernel $\Gamma^N : \measures \to \mathcal{M}(\IR^d \times [0,N))$
so that for $\eta \in \measures$, if $\hat g(x) = \int_0^N g(x, u) du / N$, then
\begin{align*}
    F^N_g(\eta)
    &:=
    \int f(\lp) \Gamma^N(\eta, d\lp) \\
    &=
    \exp\left(
        N \left\langle
            \log \frac{1}{N} \int_0^N g(x, u) du, \eta(dx)
        \right\rangle
    \right) \\
    &=
    \exp\left(
        N \left\langle \log \hat g(x) \eta(dx) \right\rangle
    \right) ,
\end{align*}
i.e., $\Gamma^N$ assigns independent labels on $[0, N]$ to each of the points in $\eta$.
It follows from Lemma~\ref{def: MP definition of limit} that for test functions of this form
\comment{(beware the different definition of $f$ to that lemma)}
the generator of $\eta^N_t$ is
\begin{align}
    \Pgen^N F^N_g(\eta)
    &=
    F^N_g(\eta)
    \theta \bigg\langle
        \gamma(x, \eta) \int r(z, \eta) \left( \hat g(z) - 1 \right) q_\theta(x, dz)
        +
        \mu_\theta(x, \eta) \left( \frac{1}{\hat g(x)} - 1 \right)
        ,
        \eta(dx)
    \bigg\rangle
\end{align}
The generator of $\lp^N_t$ is $A^N$, defined in equation~\label{eqn:lookdown_generator}.

@@

>>>>

\begin{lemma}
    \label{def: MP definition of limit}
For $G \in \mathcal{C}^{\infty}(\IR)$ with $\|G'''\|_\infty<\infty$,
and 
$f \in \mathcal{C}_{b}^\infty(\overline{\IR}^d)$, define the 
function $G_f$ by 
$G_f(\eta):=G (\langle f, \eta \rangle)$.
Let $\Pgen^N$ be the generator given by
\begin{equation}
\begin{aligned}
    \Pgen^N G_f(\eta)
    &:=
    \theta N \bigg\langle
    \gamma(x, \eta)
        \int 
        \left(G(\langle f, \eta \rangle + f(z)/N) - G(\langle f, \eta \rangle)\right)
        r(z,\eta) q_\theta(x,dz)
    \\ & \qquad \qquad {}
    +
    \left(G(\langle f, \eta \rangle - f(x)/N) - G(\langle f, \eta \rangle)\right)
    \mu_\theta(x, \eta),
    \eta(dx) \bigg\rangle .
\end{aligned}
\end{equation}
A process $(\eta^N_t)_{t \geq 0}$ 
satisfies the martingale characterization of Definition~\ref{defn:mgale_construction}
if and only if it is a solution 
to the $(\Pgen^N, \eta_0)$-martingale problem, i.e.,
if for all such test functions 
$$
    M_t :=
    G_f(\eta^N_t) - G_f(\eta^N_0)
    - \int_{0}^{t}\Pgen^N G_f(\eta^N_s)ds
$$
is a martingale (with respect to the natural $\sigma$-field).

@@

To complete the description of the process,
each line of descent
gives birth to lines at higher levels at rate $2 (N - u) c_\theta(x, \eta)$,
and each such new line chooses a level uniformly from $[u, N]$,
a spatial location $y$ from the kernel
\begin{equation} \label{eqn:qm_defn}
    q^m(x, dy, \eta)
    =
    r(y, \eta) q(x, dy) / \int_{\IR^d} r(y, \eta) q(x, dy),
\end{equation}
and the two lines swap spatial locations with probability 1/2.

@@

\begin{proof}[of Lemma \ref{lem:test_fn_tightness}]
    Again, we use the Aldous-Rebolledo criterion.
    Tightness of $\langle f, \lp_t \rangle$ for a fixed $t$
    follows from Lemma~\ref{lem:compact containment hn},
    so we need only prove conditions analogous to
    \eqref{eqn:eta_projections_goal1}
    and \eqref{eqn:eta_projections_goal2}
    applied to the martingale representation
    of equations~\eqref{eqn:f_xi_mgale} and~\eqref{eqn:f_xi_qv}.
    Rewriting~\eqref{eqn:f_xi_mgale} with $c_\theta = c_\theta(x, \eta_s)$,
    \begin{align*}
        \langle f, \lp_t \rangle
        &=
        \langle f, \lp_0 \rangle
        + M^f_t
        + \int_0^t \Big\langle
            c_\theta \int_u^N \int (f(y, u_1) + f(x, u_1)) q^m(x,dy,\eta) du_1
        \\ {} &\qquad
            + c_\theta (N - u) \int (f(y, u) - f(x, u)) q^m(x,dy,\eta)
            + (c_\theta u^2 - b_\theta u) \frac{d}{du} f,
            \lp_s
        \Big\rangle ds
        \\ &\le
        \langle f, \lp_0 \rangle
        + M^f_t
        + \int_0^t \Big\langle
            c_\theta \int_u^N \int (f(y, u_1) + f(x, u_1)) q^m(x,dy,\eta) du_1
        \\ {} &\qquad
            + c_\theta (N - u) \int (f(y, u) - f(x, u)) q^m(x,dy,\eta)
            + (c_\theta u^2 - b_\theta u) \frac{d}{du} f,
            \lp_s
        \Big\rangle ds
    \end{align*}
    The bounds analogous to \eqref{eqn:eta_projections_goal1} and \eqref{eqn:eta_projections_goal2}
    follow as in the proof of Lemma~\ref{lem:eta_projections_tightness}:
    for instance, observe that using that $c_\theta \le C_0$ for some $C_0$,
    the predictable part of this semimartingale decomposition is bounded by
    \begin{align*}
        \Big\langle
            2 C_0 \|f\|_\infty u_f
            + (1 - u/N) \gamma B^\theta_f
            + (C_0 u^2 - b_\theta u) \frac{d}{du} f,
            \lp_s
        \Big\rangle
        &\le
        \langle
            C_0 u_f^2 + \sup_x |b_\theta(x, \eta_s)| u_f \|\frac{d}{du} f\|_\infty
        \rangle ,
    \end{align*}
    which can be bounded as we did for \eqref{eqn:eta_projections_goal1}.
\end{proof}

@@

In this section we verify one of the conditions of the Markov Mapping Theorem,
namely, that ``integrating out levels'' in the generator of the lookdown process
we obtain the generator of the projected process.
We will work with test functions of the form
$$  f(\lp) = \prod_{(x, u) \in \lp} g(x, u) = \exp\left(\langle \log g, \lp \rangle \right) . $$
Furthermore, recall that $\kappa^N(\lp)(\cdot) = \lp(\cdot \times [0, N)) / N$
is the ``spatial projection operator'',
and define the probability kernel $\Gamma^N : \measures \to \mathcal{M}(\IR^d \times [0,N))$
so that for $\eta \in \measures$, if $\hat g(x) = \int_0^N g(x, u) du / N$, then
\begin{align*}
    F^N_g(\eta)
    &:=
    \int f(\lp) \Gamma^N(\eta, d\lp) \\
    &=
    \exp\left(
        N \left\langle
            \log \frac{1}{N} \int_0^N g(x, u) du, \eta(dx)
        \right\rangle
    \right) \\
    &=
    \exp\left(
        N \left\langle \log \hat g(x), \eta(dx) \right\rangle
    \right) ,
\end{align*}
i.e., $\Gamma^N$ assigns independent labels on $[0, N]$ to each of the points in $\eta$.
It follows from Lemma~\ref{def: MP definition of limit} that for test functions of this form
\comment{(beware the different definition of $f$ to that lemma)}
the generator of $\eta^N_t$ is
\begin{align} \label{eqn:pgen_defn_again}
    \begin{split}
    \Pgen^N F^N_g(\eta)
    =
    F^N_g(\eta)
        N \theta & \bigg\langle
        \gamma(x, \eta) \int r(z, \eta) \left( \hat g(z) - 1 \right) q_\theta(x, dz)
    \\ &\qquad \qquad {}
        +
        \mu_\theta(x, \eta) \left( \frac{1}{\hat g(x)} - 1 \right)
        ,
        \eta(dx)
    \bigg\rangle
    \end{split}
\end{align}
The generator of $\lp^N_t$ is $A^N$, defined in equation~\eqref{eqn:lookdown_generator}.

@@

\begin{proof}[of Lemma~\ref{lem:averaged_generators}]
First, break the generator $A^N$ into three parts, 
\begin{align*}
    A^N_1f(\lp)
    &=
    f(\lp) \sum_{(x,u)\in\lp} 2 c_\theta(x,\eta)
    \int_u^{N}\Bigg(
        \frac{1}{2}\frac{g(x,v_1)}{g(x,u)}
        \int_{\IR^d} (g(y,u)-g(x,u))q^m_{\theta}(x,dy,\eta)
    \Bigg)dv_1 ,
    \\
    A^N_2f(\lp)
    &=
    f(\lp) \sum_{(x,u)\in\lp} 2 c_\theta(x,\eta)
    \int_u^{N}\Bigg(
        \frac{1}{2}\int_{\IR^d}
            \left(\frac{g(y,v_1) + g(x,v_1)}{2} - 1\right)
        q^m_{\theta}(x,dy,\eta)
    \Bigg)dv_1 ,
    \\
    A^N_3f(\lp)
    &=
    f(\lp) \sum_{(x,u)\in\lp}\,
    \left( c_\theta(x,\eta) u^2 - b_{\theta}(x,\eta)u \right) \frac{\partial_u g(x,u)}{g(x,u)},
\end{align*}
where $q^m$ was defined in equation~\eqref{eqn:qm_defn},
so that 
$$ A^Nf(\lp) = A^N_1f(\lp) + A^N_2f(\lp) + A^N_3f(\lp). $$

We now integrate each piece against $\Gamma^N$.
First note that by the product form of $f$,
$$
    \int f(\lp) \sum_{(x,u)\in\lp} \frac{h(x,u)}{g(x,u)} \Gamma^N(\eta,d\lp)
    =
    F_g^N(\eta) \bigg\langle
        \frac{1}{N \hat g(x)} \int_0^N h(x,u) du,
        N \eta(dx)
    \bigg\rangle .
$$
Therefore,
\footnotesize
\begin{align*}
&
    \int A^N_1f(\lp) \Gamma^N(\eta,d\lp) 
\\ &\qquad =
    F_g^N(\eta) \bigg\langle \frac{c_\theta(x,\eta)}{\hat g(x)} \int_0^N \Bigg\{
        \int_u^N \Bigg(
            \frac{1}{2} g(x,v_1) \int_{\IR^d}(g(y,u)-g(x,u))q^m_{\theta}(x,dy,\eta)
        \Bigg) dv_1
    \Bigg\} du, \eta(dx) \bigg \rangle
\\ &\qquad =
    F_g^N(\eta) \bigg\langle \frac{c_\theta(x,\eta)}{\hat g(x)} \Bigg\{
    \int_{\IR^d} \left\{
        \int_{0}^{N} \int_u^N g(x,v_1) (g(y,u)-g(x,u)) dv_1 du
    \right\} q^m_{\theta}(x,dy,\eta)
    , \eta(dx) \bigg \rangle .
\end{align*}
\normalsize
For the second generator, we have
\footnotesize
\begin{align*}
&
    \int A^N_2 f(\lp) \Gamma^N(\eta, d\lp)
\\ &\qquad =
    F_g^N(\eta) \bigg\langle \frac{c_\theta(x,\eta)}{\hat g(x)} \int_0^N g(x, u) \Bigg\{
        \int_u^N\Bigg(\int_{\IR^d}
        \left(\frac{g(y,v_1)+g(x,v_1)}{2}-1\right) q^m_{\theta}(x,dy,\eta)
        \Bigg)dv_1 \Bigg\}
    , \eta(dx) \bigg\rangle
\\ &\qquad =
    F_g^N(\eta) \bigg\langle \frac{c_\theta(x,\eta)}{\hat g(x)}
    \int_{\IR^d}\Bigg\{ \int_{0}^{N}
    \int_u^N g(x,u) \left(
        g(y,v_1)+g(x,v_1)-2
    \right) dv_1 du\Bigg\} q^m_{\theta}(x,dy,\eta)
    , \eta(dx) \bigg\rangle .
\end{align*}  
\normalsize
For the third generator we have that, since $g(x,N) = 1$,
\footnotesize
\begin{align*}
&
    \int A^N_3f(\xi)\Gamma^N(\eta,d\lp)
\\ & \qquad =
    F_g^N(\eta) \bigg\langle \frac{1}{\hat g(x)}
    \int_0^N
        \left( c_\theta(x,\eta) u^2 - b_\theta(x,\eta) u \right) \partial_u g(x,u)
    du
    , \eta(dx) \bigg\rangle 
\\ & \qquad =
    F_g^N(\eta) \bigg\langle \frac{1}{\hat g(x)}
    \int_0^N
        \left( b_\theta(x,\eta) - 2 c_\theta(x,\eta) u \right) (g(x,u) - 1)
    du
    , \eta(dx) \bigg\rangle .
\end{align*}
\normalsize

Note that $2 \int_0^N \int_u^N g(x,v_1) g(y,u) dv_1 du = N^2 \hat g(x) \hat g(y)$,
and so
\footnotesize
\begin{align*}
&
    \int_{0}^{N} \int_u^N g(x,v_1) (g(y,u)-g(x,u)) dv_1 du
    + \int_{0}^{N} \int_u^N g(x,u) \left(g(y,v_1)+g(x,v_1)-2\right)dv_1 du
\\ &\qquad = 
    N^2 \hat{g}(x)(\hat{g}(y) - 2) + 2\int_{0}^{N} u g(x,u) du.
\end{align*}
\normalsize

Combining the last equations,
and using the fact that
$N c_\theta(x,\eta) - b_\theta(x,\eta) = \theta \mu_\theta(x,\eta)$,
we have 
\begin{align*}
&
    \int \Bigg( A^N_1 f(\xi) + A^N_2 f(\xi) + A^N_3 f(\xi) \Bigg) \Gamma^N(\eta,d\lp)
\\ &\qquad =
    F_g^N(\eta) \bigg\langle
        c_\theta(x,\eta) N^2 \int_{\IR^d} (\hat g(y) - 2) q^m(x,dy,\eta)
        +
        \frac{1}{\hat g(x)} c_\theta(x,\eta) \int_0^N 2u du
\\ &\qquad \qquad \qquad {}
        +
        \frac{1}{\hat g(x)} b_\theta(x,\eta) \int_0^N (g(x,u) - 1) du
    , \eta(dx) \bigg\rangle 
\\ &\qquad =
    F_g^N(\eta) \bigg\langle
        c_\theta(x,\eta) N^2 \int_{\IR^d} (\hat g(y) - 1) q^m(x,dy,\eta)
        +
        N^2 c_\theta(x,\eta) \left(\frac{1}{\hat g(x)} - 1\right)
\\ &\qquad \qquad \qquad {}
        +
        N b_\theta(x,\eta) \left(1 - \frac{1}{\hat g(x)}\right)
    , \eta(dx) \bigg\rangle 
\\ &\qquad =
    F_g^N(\eta) N \bigg\langle
        N c_\theta(x,\eta) \int_{\IR^d} (\hat g(y) - 1) q^m(x,dy,\eta)
        +
        \theta \mu_\theta(x,\eta) \left(\frac{1}{\hat g(x)} - 1\right)
    , \eta(dx) \bigg\rangle  .
\end{align*}
This matches equation~\eqref{eqn:pgen_defn_again}, as desired,
because $N c_\theta(x,\eta) q_m(x,dy,\eta) = \theta \gamma(x,\eta) q_\theta(x,dy)$.
\end{proof}
